---
title: "MattHill"
author: "Samuel Keat"
date: "27 October 2019"
output: html_document
---

## If "rlang" must be loaded manually prior to installation

```{r}
library("rlang", lib.loc="~/R/win-library/3.6")
```


## Set the working directory

```{r}
##setwd("C:/Users/Sam/Desktop/Voluntary Projects/Matt Hill/R/bin")
```

## Step 0.1 - Installing the packages and relevant dependencies using BioConductor

```{r}
library(readr)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::version()

## Required
BiocManager::install(c("AUCell", "RcisTarget"))
BiocManager::install(c("GENIE3")) # Optional. Can be replaced by GRNBoost

## Optional (but highly recommended):
# To score the network on cells (i.e. run AUCell):
BiocManager::install(c("zoo", "mixtools", "rbokeh"))
# For various visualizations and perform t-SNEs:
BiocManager::install(c("DT", "NMF", "pheatmap", "R2HTML", "Rtsne"))
# To support paralell execution (not available in Windows):
BiocManager::install(c("doMC", "doRNG"))
# To export/visualize in http://scope.aertslab.org
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("aertslab/SCopeLoomR", build_vignettes = TRUE, force = TRUE)
```

## Step 0.2 - Check versions of installed packages

Ensure versions of packages are: AUCell >=1.4.1 (minimum 1.2.4), RcisTarget>=1.2.0 (minimum 1.0.2) and GENIE3>=1.4.0 (minimum 1.2.1).

If not, re-run the previous step.

```{r}
packageVersion("AUCell")
packageVersion("RcisTarget")
packageVersion("GENIE3")
```

## Step 0.3 - Finally, install SCENIC

Select option 1 to install and update ALL packages and dependencies.

```{r}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("aertslab/SCENIC",force=TRUE)
library(SCENIC)
packageVersion("SCENIC")
```

## Step 1 - Convert the Human Transcription Factors to Mouse

Conversion using 'gorth' from the package 'gProfileR'. First, install 'gProfileR' and load library.

```{r}
install.packages("gProfileR")
library(gProfileR)

##install.packages("gprofiler2")
##library(gprofiler2)
```

Then perform the conversion using the 'gorth' function.

```{r}
humTF <- read.table("C:/Users/Sam/Desktop/Voluntary Projects/Matt Hill/R/resources/TFs_Ensembl_v_1.01.txt", quote="\"", comment.char="")

mouseTF=gorth(query=as.array(humTF$V1),source_organism="hsapiens",target_organism="mmusculus")
```

Save this list as a .txt file

```{r}
write.table(mouseTF$target.ensg,file="C:/Users/Sam/Desktop/Voluntary Projects/Matt Hill/R/resources/MM_TFs_Ensembl.txt",quote=FALSE,eol="\n",row.names=FALSE,col.names=F)
```

Save whole dataframe as a .csv

```{r}
write.csv(mouseTF,file="C:/Users/Sam/Desktop/Voluntary Projects/Matt Hill/R/resources/MM_TFs_full.csv",quote=TRUE,eol="\n",row.names=FALSE)
```

## Step 2 - Uploading and cleaning the data using SCENIC

Upload and format the GEO dataset.

```{r}
if (!requireNamespace("GEOquery", quietly = TRUE)) BiocManager::install("GEOquery")
library(GEOquery)
geoFile <- getGEOSuppFiles("GSE123025", makeDirectory=FALSE)
gzFile <- grep("data", basename(rownames(geoFile)), value=TRUE)
txtFile <- gsub(".gz", "", gzFile)
gunzip(gzFile, destname=txtFile, remove=TRUE, overwrite=TRUE)

library(data.table)
geoData <- fread(txtFile, sep=",")
geneNames <- unname(unlist(geoData[,1, with=FALSE]))
exprMatrix <- as.matrix(geoData[,-1, with=FALSE])
rm(geoData)
dim(exprMatrix)
rownames(exprMatrix) <- geneNames
exprMatrix <- exprMatrix[unique(rownames(exprMatrix)),]
exprMatrix[1:5,1:4]

file.remove(txtFile)
```

Load the cell types from the AUCell package.

```{r}
cellLabels <- paste(file.path(system.file('examples', package='AUCell')), "mouseBrain_cellLabels.tsv", sep="/")
cellLabels <- read.table(cellLabels, row.names=1, header=TRUE, sep="\t")
cellInfo <- as.data.frame(cellLabels)
colnames(cellInfo) <- "Class"
```

Make sure all the required packages are loaded.

```{r}
library(GENIE3)
library(AUCell)
library(RcisTarget)
library(SCENIC)
```

Save this information into a .loom file for further analysis.

```{r}
library(SCopeLoomR)
dir.create("data")
loom <- build_loom("data/GSE123025.loom", dgem=exprMatrix)
library(SCENIC)
close_loom(loom)
```

## Step 3 - Running the SCENIC workflow.

Load in the data.

```{r}
loomPath <- system.file(package="SCENIC", "examples/mouseBrain_toy.loom")
library(SCopeLoomR)
loom <- open_loom(loomPath, mode="r")
exprMat <- get_dgem(loom)
cellInfo <- get_cellAnnotation(loom)
close_loom(loom)

dim(exprMat)
```

Create a directory to store the workings.

```{r}
knitr::opts_knit$set(root.dir = 'example_results/SCENIC_MouseBrain')
```

Head the known cell data.

```{r}
head(cellInfo)
```

Count the number of cell types and store in a seperate table.

```{r}
cellInfo <- data.frame(cellInfo)
cellTypeColumn <- "Class"
colnames(cellInfo)[which(colnames(cellInfo)==cellTypeColumn)] <- "CellType"
cbind(table(cellInfo$CellType))
```

Create a directory to store this information.

```{r}
dir.create("int")
```

Save the information about the cells into the newly created directory in R Data Single (RDS) format.

```{r}
saveRDS(cellInfo, file="int/cellInfo.Rds")
```

Assign each of the variables as colours for later use in a heatmap.

```{r}
colVars <- list(CellType=c("microglia"="forestgreen", 
                           "endothelial-mural"="darkorange", 
                           "astrocytes_ependymal"="magenta4", 
                           "oligodendrocytes"="hotpink", 
                           "interneurons"="red3", 
                           "pyramidal CA1"="skyblue", 
                           "pyramidal SS"="darkblue"))
colVars$CellType <- colVars$CellType[intersect(names(colVars$CellType), cellInfo$CellType)]
saveRDS(colVars, file="int/colVars.Rds")
plot.new(); legend(0,1, fill=colVars$CellType, legend=names(colVars$CellType))
```

First, check if the .feather files are available. If not, download them (they may need to be downloaded manually if the settings initilisation causes the R session to terminate)

```{r}
library(SCENIC)

dbFiles <- c("https://resources.aertslab.org/cistarget/databases/mus_musculus/mm9/refseq_r45/mc9nr/gene_based/mm9-500bp-upstream-7species.mc9nr.feather","https://resources.aertslab.org/cistarget/databases/mus_musculus/mm9/refseq_r45/mc9nr/gene_based/mm9-tss-centered-10kb-7species.mc9nr.feather")

## File names after download
downloadedDBFiles <- c("cisTarget_databases/mm9-500bp-upstream-7species.mc9nr.feather","cisTarget_databases/mm9-tss-centered-10kb-7species.mc9nr.feather")


## The databases are downloaded only in the case when the files do not exist.
if( !(file.exists(downloadedDBFiles[1]) & file.exists(downloadedDBFiles[2])) ){
  
  ## Create and move to the new folder
  dir.create("cisTarget_databases");
  setwd("cisTarget_databases") 
  for(featherURL in dbFiles)
  {
    download.file(featherURL, destfile=basename(featherURL)) # saved in current dir
    descrURL <- gsub(".feather$", ".descr", featherURL)
    if(file.exists(descrURL)) download.file(descrURL, destfile=basename(descrURL))
  }
  setwd("..")
  
}
```

(gene filtering by biologically relevant cells of the .feather)

Initialise settings for SCENIC before operation, selecting mouse (mgi) as organism.

```{r}
org="mgi"
dbDir="cisTarget_databases"
myDatasetTitle="SCENIC Mouse brain - GSE123025"
data(defaultDbNames)
dbs <- defaultDbNames[[org]]
scenicOptions <- initializeScenic(org=org, dbDir=dbDir, dbs=dbs, datasetTitle=myDatasetTitle, nCores=10) 
```

## Step 5 - Running the GENIE3 Network Inference

Build up a co-expression network, filtering out certain genes using the SCENIC options defined previously.

```{r}
# Filter out genes expressed at very low levels (possible noise) or in too few cells
# Only genes in the RcisTarget database are kept too
genesKept <- geneFiltering(exprMatrix, scenicOptions, minCountsPerGene=3*.01*ncol(exprMatrix), minSamples=ncol(exprMatrix)*.01)

# Check how many of the transcription factors have been removed
interestingGenes <- as.vector(mouseTF$target.name)

# Any ones that are missing are stored as an .RDS file
interestingGenes[which(!interestingGenes %in% toupper(genesKept))]
```

Now filter out the co-expression matrix to only contain these genes.

```{r}
exprMat_filtered <- exprMatrix[genesKept, ]
# Then find out the dimensions of this new expression matrix
dim(exprMat_filtered)
```

Now remove the old expression matrix.

```{r}
rm(exprMatrix)
```

In order to determine positive or negative associations (activation or repression), the targets are split into positively and negatively correlated targets.

```{r}
runCorrelation(exprMat_filtered, scenicOptions)
```

Then, run the GENIE3 network inference, first normalising the data (if it hasn't already been) and then running under SCENIC settings

```{r}
exprMat_filtered_log <- log2(exprMat_filtered+1) 
#runGenie3(exprMat_filtered_log, scenicOptions)
```

Build-up and provide scores for the Gene Regulatory Network (GRN)

```{r}
exprMat_log <- log2(exprMat+1)
scenicOptions@settings$dbs <- scenicOptions@settings$dbs["10kb"] # Toy run settings
runSCENIC_1_coexNetwork2modules(scenicOptions)
runSCENIC_2_createRegulons(scenicOptions,coexMethod=c("top5perTarget")) # Toy run settings
runSCENIC_3_scoreCells(scenicOptions, exprMat_log)
```

Export this GRN and scores to scope.

```{r}
export2scope(scenicOptions, exprMatrix)
```

## Binarise the regulon activity (step 4)

Set the input values, plotting all available figures

```{r}
skipBoxplot=FALSE
skipHeatmaps=FALSE
skipTsne=FALSE
```

Assign the cells and save as a binary matrix

```{r}
nCores <- getSettings(scenicOptions, "nCores")
regulonAUC <- loadInt(scenicOptions, "aucell_regulonAUC")
thresholds <- loadInt(scenicOptions, "aucell_thresholds")
thresholds <- getThresholdSelected(thresholds)

# Assign cells
regulonsCells <- setNames(lapply(names(thresholds), 
                                 function(x) {
                                   trh <- thresholds[x]
                                   names(which(getAUC(regulonAUC)[x,]>trh))
                                 }),names(thresholds))
### Convert to matrix (regulons with zero assigned cells are lost)
regulonActivity <- reshape2::melt(regulonsCells)
binaryRegulonActivity <- t(table(regulonActivity[,1], regulonActivity[,2]))
class(binaryRegulonActivity) <- "matrix"
saveRDS(binaryRegulonActivity, file=getIntName(scenicOptions, "aucell_binary_full"))
```

Plot heatmap

```{r}
if(!skipHeatmaps)
{
  regulonSelection <- loadInt(scenicOptions, "aucell_regulonSelection", ifNotExists="null", verbose=FALSE)
  if(is.null(regulonSelection)) 
    regulonSelection <- regulonSelections(binaryRegulonActivity, binaryRegulonActivity_nonDupl, minCells)

  cellInfo <- loadFile(scenicOptions, getDatasetInfo(scenicOptions, "cellInfo"), ifNotExists="null")
  cellInfo <- data.frame(cellInfo)
  colVars <- loadFile(scenicOptions, getDatasetInfo(scenicOptions, "colVars"), ifNotExists="null")


  ### Plot heatmap:
  for(selRegs in names(regulonSelection$labels))
  {
    if(length(regulonSelection[[selRegs]])>1)
    {
      regulonSelection[[selRegs]] <- regulonSelection[[selRegs]][which(regulonSelection[[selRegs]] %in% rownames(binaryRegulonActivity))]
      binaryMat <- binaryRegulonActivity[regulonSelection[[selRegs]],,drop=FALSE]

      fileName <- paste0(getOutName(scenicOptions, "s4_binaryActivityHeatmap"),selRegs)

      fileName <- .openDevHeatmap(fileName=fileName, devType=getSettings(scenicOptions, "devType"))
        NMF::aheatmap(binaryMat, scale="none", revC=TRUE, main=selRegs,   
                      annCol=cellInfo[colnames(binaryMat),, drop=FALSE],
                      annColor=colVars,
                      color = c("white", "black"),
                      filename=fileName)
      if(getSettings(scenicOptions, "devType")!="pdf") dev.off()
    }
  }
}  
t-SNE on binary activity (optional)
if(!skipTsne)
{
  tSNE_fileName <- tsneAUC(scenicOptions, aucType="Binary", filePrefix=getIntName(scenicOptions, "tsne_prefix"), onlyHighConf=FALSE) # default: nPcs, perpl, seed
  tSNE <- readRDS(tSNE_fileName)

  # AUCell (activity) as html: 
  fileName <- getOutName(scenicOptions, "s4_binarytSNE_colAct")
  plotTsne_regulonActivityHTML(scenicOptions, exprMat, fileName, tSNE) #open the resulting html locally

  # Plot cell properties:
  sub <- ""; if("type" %in% names(tSNE)) sub <- paste0("t-SNE on ", tSNE$type)
  cellInfo <- loadFile(scenicOptions, getDatasetInfo(scenicOptions, "cellInfo"), ifNotExists="null")
  colVars <- loadFile(scenicOptions, getDatasetInfo(scenicOptions, "colVars"), ifNotExists="null")
  pdf(paste0(getOutName(scenicOptions, "s4_binarytSNE_colProps"),".pdf"))
  plotTsne_cellProps(tSNE$Y, cellInfo=cellInfo, colVars=colVars, cex=1, sub=sub)
  dev.off()
}
```

Plot tSNE

```{r}
if(!skipTsne)
{
  tSNE_fileName <- tsneAUC(scenicOptions, aucType="Binary", filePrefix=getIntName(scenicOptions, "tsne_prefix"), onlyHighConf=FALSE) # default: nPcs, perpl, seed
  tSNE <- readRDS(tSNE_fileName)

  # AUCell (activity) as html: 
  fileName <- getOutName(scenicOptions, "s4_binarytSNE_colAct")
  plotTsne_regulonActivityHTML(scenicOptions, exprMat, fileName, tSNE) #open the resulting html locally

  # Plot cell properties:
  sub <- ""; if("type" %in% names(tSNE)) sub <- paste0("t-SNE on ", tSNE$type)
  cellInfo <- loadFile(scenicOptions, getDatasetInfo(scenicOptions, "cellInfo"), ifNotExists="null")
  colVars <- loadFile(scenicOptions, getDatasetInfo(scenicOptions, "colVars"), ifNotExists="null")
  pdf(paste0(getOutName(scenicOptions, "s4_binarytSNE_colProps"),".pdf"))
  plotTsne_cellProps(tSNE$Y, cellInfo=cellInfo, colVars=colVars, cex=1, sub=sub)
  dev.off()
}

```

Quick

```{r}
runSCENIC_4_aucell_binarize(scenicOptions,skipBoxplot=FALSE,skipHeatmaps=FALSE,skipTsne=FALSE,exprMat=NULL)
```

## Extra Steps

Installing and loading RDAVIDWebService.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RDAVIDWebService")
```

Load RDAVIDWebService

Optional: If Java isn't working (rJava won't load), install Java and set the system environment to the Java directory (e.g. Sys.setenv(JAVA_HOME="C:/Program Files/Java/jre1.8.0_241"))

```{r}
Sys.setenv(JAVA_HOME="C:/Program Files/Java/jre1.8.0_241")
library(rJava)
library(RDAVIDWebService)
```

Running DAVID

```{r}
RDAVIDWebService::genes(interestingGenes)
```

Load tidyverse and dplyr.

```{r}
library(tidyverse)
library(dplyr)
```

Running GProfileR gene enrichment finding tool, first to find the biological process (BP).

```{r}
BPresults=gprofiler(interestingGenes, organism="mmusculus", min_set_size=5, min_isect_size=5, correction_method="fdr", src_filter="GO:BP")
BPlist=BPresults$term.name
BPgene=BPresults$intersection
BPannotation=bind_cols(data.frame(BPgene),data.frame(BPlist))
```

Then to find the molecular function (MF).

```{r}
MFresults=gprofiler(interestingGenes, organism="mmusculus", min_set_size=5, min_isect_size=5, correction_method="fdr", src_filter="GO:MF")
MFlist=MFresults$term.name
MFgene=MFresults$intersection
MFannotation=bind_cols(data.frame(MFgene),data.frame(MFlist))
```

Running with CC

```{r}
CCresults=gprofiler(interestingGenes, organism="mmusculus", min_set_size=5, min_isect_size=5, correction_method="fdr", src_filter="GO:CC")
CClist=CCresults$term.name
CCgene=CCresults$intersection
CCannotation=bind_cols(data.frame(CCgene),data.frame(CClist))
```